---
output:
  pdf_document: default
  html_document: default
---
# Future Emissions
Add text here

## Install the kayadata R-package
* In the lower-left panel you find your console where you can execute commands. 
* Your first task is to set up your computational environment using the command ```.libPaths()```
* In the Console, type the command below using the directory of your project. Note that you need to use ```/``` rather than ```\```.

```{r, eval = FALSE}
# Set your working directory:
setwd("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430")

# Tell R where to store your packages:
.libPaths("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/renv")

# Install the kayadata R-package
install.packages("kayadata")

```

* You will be asked to install a wide range of dependencies
* Type ```Y``` when asked whether you want to proceed
```{r, eval = FALSE}
Do you want to proceed? [Y/n]: Y
```


## Using the kayadata package

```{r, eval = FALSE}
library(kayadata)

# Read data for your country of choice
canada_data <- get_kaya_data("Canada") 
print(canada_data)
```

* To find our what the different columns are, read the documentation of the ```get_kaya_data``` function using ```?get_kaya_data```

```{r, eval = FALSE}
?get_kaya_data
```

* Let's make a few plots

```{r, eval = FALSE}
plot_kaya(canada_data, "P", y_lab = "Population (billions)")
plot_kaya(canada_data, "G", y_lab = "GDP (trillions of constant 2015 USD)")
plot_kaya(canada_data, "E", y_lab = "Total primary energy consumption (quads)")
plot_kaya(canada_data, "ef", y_lab = "CO2 emissions from fossil fuel consumption (millions of metric tons)")
```

* Let's have a look at the projection for the year 2050

```{r, eval = FALSE}
canada_2050 <- project_top_down("Canada", 2050)
print(canada_2050)
```



