---
output:
  html_document: default
  pdf_document: default
---
# Climate Model Analysis (I)

## Earth System Grid Federation (ESGF)

* Visit the [ESGF website](https://esgf.llnl.gov/)
* Click on [Metagrid web application](https://aims2.llnl.gov/search)
* We want to download monthly precipitation simulated by the Candian Earth System Model for the historical period
* To find the file, you will need filter the data base by adding key words
* In the keyword section, add ```CMIP``` (Coupled Model Intercomparison Project), ```CCCma``` (Canadian Centre for Climate Modelling and Analysis), ```CanESM5``` (Canadian Earth System Model version 5), ```historical``` (historical period), ```r10i1p1f1``` (ensemble member ID), ```Amon``` (realm: atmosphere, time resolution: monthly), and ```pr.gn``` (precipitation)
* Once you find the file, download the corresponding wget script
* To find different files, clear keywords and re-enter new keywords:

```
# Monthly precipitation
CMIP6.CMIP.CCCma.CanESM5.historical.r10i1p1f1.Amon.pr.gn
CMIP6.ScenarioMIP.CCCma.CanESM5.ssp585.r10i1p1f1.Amon.pr.gn

# Daily precipitation
CMIP6.CMIP.CCCma.CanESM5.historical.r10i1p1f1.day.pr.gn
CMIP6.ScenarioMIP.CCCma.CanESM5.ssp585.r10i1p1f1.day.pr.gn
```

* Variable names are listed [here](https://pcmdi.llnl.gov/mips/cmip3/variableList.html)
* For example, search for ```precipitation``` (Windows search: ```Ctrl + F```, Mac search: ```Command + F```), you will find the variable description, ID, and physical units ```precipitation_flux, pr, kg m-2 s-1```

* Open one of the wget files using a text editor and search for something that looks like this:

```
#These are the embedded files to be downloaded
download_files="$(cat <<EOF--dataset.file.url.chksum_type.chksum
'pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc' 'http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/pr/gn/v20190429/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc' 'SHA256' '2eb2e7fa943902a9fd4f726a98e6ba5713b23558a8da5bda8c992ef374aca6d4'
EOF--dataset.file.url.chksum_type.chksum
)"
```

There you find the URL to the file that you want to download (it may look slightly different for you):

```
http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/pr/gn/v20190429/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc
```

* Copy-paste the URL to another text file
* Repeat the same for all the other files you downloaded
* Additional information: You can also first add all your files to a cart and then download a single wget file that contains all URLs

* To download the netcdf files use the ```download.file``` function, e.g.:

```{r, eval = FALSE}
# Specify the URL of the file you want to download
url <- "http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/pr/gn/v20190429/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc"

# Specify the file name and location where you want to save the file on your computer
file_name <- "pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc"
file_path <- "C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/data/"

# Call the download.file() function, passing in the URL and file name/location as arguments
download.file(url, paste(file_path, file_name, sep = ""), mode = "wb")
```

* Note that ```file_path``` should contain ```/``` rather than ```\``` and may not contain white space or special characters (e.g. ```C:\Users\seile\OneDrive - Queen's University\Documents``` will not work as a directory)

## Open, process, and plot netcdf files

```{r}
library(terra)
rm(list = ls())
data <- rast("data/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
print(data)
```
* You see the spatial dimensions, spatial resolution, spatial extent in longitude and latitude, variable name, etc.
* Each layer is one time step, starting in Jan 1850
* Let's calculate the mean value over all time steps and plot the data

```{r}
library(maps)
data <- terra::mean(data) # annual mean
terra::plot(data)
map("world2", add = TRUE)
```

* The ```terra``` package comes with different colors (```?map.pal```), e.g.

```{r}
my.col = map.pal("blues", n=100)
terra::plot(data, col = my.col)
map("world2", add = TRUE)
```

* Let's plot how precipitation is projected to change between the 1995-2014 and 2081-2100 periods
* Download future precipitation projections:

```{r, eval = FALSE}
# Get the URL for the SSP5-8.5 scenario
url <- "http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgD_dataroot/AR6/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/Amon/pr/gn/v20190429/pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc"

# Specify the file name and location where you want to save the file on your computer
file_name <- "pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc"
file_path <- "C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/data/"

# Call the download.file() function, passing in the URL and file name/location as arguments
download.file(url, paste(file_path, file_name, sep = ""), mode = "wb")
```

* Load, process, and plot the data

```{r}
library(terra)

# Read in data
hist <- rast("data/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
ssp5 <- rast("data/pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")

# Subset last 20 years from both files
n <- nlyr(hist) # number of layers = number of time steps
hist <- subset(hist, (n - 239):n) # last 240 months = 20 years

n <- nlyr(ssp5) # number of layers = number of time steps
ssp5 <- subset(ssp5, (n - 239):n) # last 240 months = 20 years

delta <- mean(ssp5) - mean(hist)

# Let's change the unit of precipitation from kg m2 s-1 to kg m2 day-1
delta <- delta * 86400

my.col = map.pal("differences", n = 100)
plot(delta, col = my.col)
map("world2", add = TRUE)
```

* Nice, but we want to have the divergence from red to blue at zero
* To achieve this, we need to define the breaks of the color palette
* Also, we will reverse the color, so that blue implies a projected increase in precipitation

```{r}
# Set the desired range for the palette
min_val <- -8
max_val <- 8

# Define breaks to control the min and max of the palette
breaks <- seq(min_val, max_val, 2)

my.col <- rev(map.pal("differences", n = length(breaks) - 1))

plot(delta,
     type = "continuous",
     breaks = breaks,
     col = my.col)
map("world2", add = TRUE)
```

* Great, but are the projected changes statistically significant?
* Let's apply a t-test for each grid cell and mark all grid cells where the projected change is statistically significant at the 0.05-level
* Before we do that, let's practice how to do a t-test in the first place

```{r}
# Create two random data sets
a <- runif(n = 100, min = 10, max = 20)
b <- runif(n = 100, min = 11, max = 21)

# Test whether the mean of both data sets is statistically different
pvalue <- t.test(x = a,
                 y = b,
                 alternative = c("two.sided"))$p.value
print(pvalue)
```

* OK, let's apply this to the projected change in precipitation

```{r}
# Load data
hist <- rast("data/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
ssp5 <- rast("data/pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")

# Define a function that performs a t-test and returns the p-value
ttest.fun <- function(x) {
  n <- length(x) / 2
  a <- x[1:n]
  b <- x[(n + 1):(2 * n)]
  pvalue <- t.test(x = a,
                   y = b,
                   alternative = c("two.sided"))$p.value
  return(pvalue)
}

# Apply the t-test function to each cell in the raster stacks
pvalue.gridCell <- app(c(hist, ssp5), fun = ttest.fun)

# Plot the p-values raster
plot(pvalue.gridCell)
```

* Now let's create a mask where all grid cells with statistically significant changes have no value (NA) while all others have a value of one

```{r}
pvalue.gridCell[pvalue.gridCell<0.05] <- NA # statistically significant changes = 0
mask <- pvalue.gridCell - pvalue.gridCell + 1 # nost statistically significant changes = 1
```

* Great, now let's plot our previous map that shows the projected change in precipitation and mask out all grid cells where the projected changes are not statistically significant

```{r}

plot(
  delta,
  type = "continuous",
  breaks = breaks,
  col = my.col,
  main = "Projected Changes in Annual Mean Precipitation \n from 1995-2014 to 2081-2100 (CanESM5, SSP5-8.5)",
  font.main = 1
)

plot(mask,
     add = TRUE,
     legend = FALSE,
     col = "white")
map("world2", add = TRUE)
mtext("mm per day", side = 4, line = 0.75)
```

## Climate Extremes
* Next we will use the [ClimInd](https://cran.r-project.org/web/packages/ClimInd/index.html) R-package to compute climate indices
* The package contains 138 standard climate indices at monthly, seasonal and annual resolution
* The indices characterize different aspects of the frequency, intensity and duration of extreme events
* Input variables consist of surface air temperature, precipitation, relative humidity, wind speed, cloudiness, solar radiation, and snow cover
* The package includes Temperature based indices (42), Precipitation based indices (22), Bioclimatic indices (21), Wind-based indices (5), Aridity/continentality indices (10), Snow-based indices (13), Cloud/radiation based indices (6), Drought indices (8), Fire indices (5), and Tourism indices (5)
* Let's first practice a simple example using random data

```{r}
library(ClimInd)

# Generate random data with dates in month/day/year format
dates <- seq(as.Date("1990/1/1"), as.Date("1999/12/31"), "days")
dates <- format(dates, "%m/%d/%Y")
data <- runif(n=length(dates), min = 0, max = 5)
names(data) <- dates

# Let's pretend our data consists of daily precipitation in mm/day
# Let's compute consecutive dry days

my.cdd <- cdd(data, data_names = NULL, time.scale = "YEAR", na.rm = FALSE)
print(my.cdd)
```

* Great, now let's apply our new skill to climate model data
* Download future daily temperature and precipitation data (CanESM5, SSP5-8.5, r10i1p1f1)

```{r, eval = FALSE}

# Daily future precipitation
file_path <- "C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/data/" # Change this path to wherever you wish to store your data

# Specify the URL of the file you want to download
url <- "http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgD_dataroot/AR6/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/day/pr/gn/v20190429/pr_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc"

# Specify the file name and location where you want to save the file on your computer
file_name <- "pr_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc"

# Download file
options(timeout=300) # If the download requires more time, then increase the number of seconds
download.file(url, paste(file_path, file_name, sep = ""), mode = "wb", timeout=300)


# Daily future temperature
# Specify the URL of the file you want to download
url <- "http://esgf-data1.llnl.gov/thredds/fileServer/css03_data/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/day/tas/gn/v20190429/tas_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc"

# Specify the file name and location where you want to save the file on your computer
file_name <- "tas_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc"

# Call the download.file() function, passing in the URL and file name/location as arguments
options(timeout=300) # If the download requires more time, then increase the number of seconds
download.file(url, paste(file_path, file_name, sep = ""), mode = "wb")
```


```{r}
library(terra)
library(ClimInd)
library(maps)

# Load daily precipitation data
pr <- rast("data/pr_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc")

# Let's subset four years for testing purposes
data <- subset(pr, 1:1460)
data <- data * 86400 # convert from kg m-2 s-1 to mm day-1

  dates <- time(data)
  dates <- format(dates, "%m/%d/%Y")
  names(data) <- dates
  
# Define a function that calculated consecutive dry days
cdd.fun <- function(x) {
   my.cdd <- cdd(data = x, time.scale = "YEAR", na.rm = FALSE)
  return(my.cdd)
}

# Apply function to each gridcell
cdd.gridCell <- app(data, fun = cdd.fun)

# You now have a raster stack with one leayer for each year
# print(cdd.gridCell)

# Let's plot the first year
my.col = rev(map.pal("magma", n=100))
firstLayer <- subset(cdd.gridCell, 1:1)
plot(firstLayer, col = my.col, main = "Consecutive Dry Days")
map("world2", add = TRUE)
```

## Consecutive Wet Days

* Compute consecutive wet days using the ```cwd``` function provided by the [ClimInd](https://cran.r-project.org/web/packages/ClimInd/index.html) R-package
* Use the script above as a template
* Your results should look similar to the Figure below

```{r, echo = FALSE}
library(terra)
library(ClimInd)
library(maps)

# Load daily precipitation data
pr <- rast("data/pr_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc")

# Let's subset four years for testing purposes
data <- subset(pr, 1:1460)
data <- data * 86400 # convert from kg m-2 s-1 to mm day-1

  dates <- time(data)
  dates <- format(dates, "%m/%d/%Y")
  names(data) <- dates
  
# Define a function that calculated consecutive dry days
cwd.fun <- function(x) {
   my.cwd <- cwd(data = x, time.scale = "YEAR", na.rm = FALSE)
  return(my.cwd)
}

# Apply function to each gridcell
cwd.gridCell <- app(data, fun = cwd.fun)

# You now have a raster stack with one leayer for each year
# print(cdd.gridCell)

# Let's plot the first year
my.col = map.pal("blues", n=100)
firstLayer <- subset(cwd.gridCell, 1:1)
plot(firstLayer, col = my.col, main = "Consecutive Wet Days")
map("world2", add = TRUE)
```

## Linear trend of consecutive dry days

* Compute the linear trend in consecutive dry days from 2015 to 2100
* Before you start, let me briefly show you how you calculate the linear trend of a variable

```{r}
# Sample time series data
time <- 1:10  # Independent variable (e.g., time in years)
temperature <- runif(n = length(time), min = 1, max = 3) * time

# Fit a linear model (trend)
trend_model <- lm(temperature ~ time)

# Extract the slope (coefficient for 'time')
slope <- coef(trend_model)["time"]

# Print the slope
print(slope)

# Plot time series
plot(time, temperature,type = "l")
abline(trend_model, col="red", lwd=2)
```

* Calculating climate extremes from daily data is computationally expensive
* To speed things up, let's focus on one region
* Let's first crop Australia and then perform the calculation for the region

```{r, eval = FALSE}
# Load daily precipitation data
data <- rast("data/pr_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc")
# Let's focus on Australia
data <- crop(x=data, y = c(100,170,-50,0))
```

* Now, calculate the trend of consecutive dry days for each grid cell
* To achieve this, define a function that calculates the slope of a linear regression
* Apply this function to each grid cell
* Your results should look similar to the Figure below

```{r}
library(terra)
library(ClimInd)
library(maps)

# Load daily precipitation data
data <- rast("data/pr_day_CanESM5_ssp585_r10i1p1f1_gn_20150101-21001231.nc")

# Let's focus on Australia
data <- crop(x=data, y = c(100,170,-50,0))

data <- data * 86400 # convert from kg m-2 s-1 to mm day-1

  dates <- time(data)
  dates <- format(dates, "%m/%d/%Y")
  names(data) <- dates
  
# Define a function that calculated consecutive dry days
cdd.fun <- function(x) {
   my.cdd <- cdd(data = x, time.scale = "YEAR", na.rm = FALSE)
  return(my.cdd)
}

# Apply function to each gridcell
cdd.gridCell <- app(data, fun = cdd.fun)

n <- nlyr(cdd.gridCell)
time <- seq(1, n, 1)

# Define a function that calculates a linear trend
trend.fun <- function(x) {
   linear.model <- lm(x ~ time)
   trend <- coef(linear.model)["time"]
  return(trend)
}

# Apply function to each gridcell
trend <- app(cdd.gridCell, fun = trend.fun)

# Set the desired range for the palette
min_val <- -1
max_val <- 1

# Define breaks to control the min and max of the palette
breaks <- seq(min_val, max_val, 0.1)

my.col <- map.pal("differences", n = length(breaks) - 1)

plot(trend, col = my.col, 
  type = "continuous",
  breaks = breaks,
  main = "Change in Consecutive Dry Days per year (2015-2100)")
map("world2", add = TRUE)
```

## Consecutive dry days time series

* Next, plot a time series of consecutive dry days for the cropped region
* To do that, use the ```global``` function in the ```terra``` package
* Your results should look like the Figure below

```{r}
cdd.mean <- global(cdd.gridCell, mean, na.rm = TRUE)

years <- rownames(cdd.mean)
plot(x = years, y = cdd.mean$mean, 
     type = "l",
     ylab = "CDD in Australia")

abline(trend_model, col="red", lwd=2)
```

## Analysis of an extreme climate index of your choice and region

* Repeat the previous two exercises but for a different region of the globe and a different climate index
* The climate indices are provided by the [ClimInd](https://cran.r-project.org/web/packages/ClimInd/index.html) package
