---
output:
  pdf_document: default
  html_document: default
---
# Climate Model Analysis (II)

Install the R package [climetrics](https://cran.r-project.org/web/packages/climetrics/index.html):

```{r, eval = FALSE, echo - TRUE}
setwd("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430")
.libPaths("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/renv")
library(renv)
renv::install("climetrics")
library("climetrics")
```

Download climate model data from ESGF by clicking on the links and save them in a new folder called ```data``` that you put inside your working directory, e.g. ```C:\Users\seile\Documents\ENSC430\data```

* [Monthly mean temperature, historical](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/tas/gn/v20190429/tas_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc)
* [Monthly min temperature, historical](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/tasmin/gn/v20190429/tasmin_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc)
* [Monthly max temperature, historical](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/tasmax/gn/v20190429/tasmax_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc)
* [Monthly mean temperature, SSP585](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgD_dataroot/AR6/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/Amon/tas/gn/v20190429/tas_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc)
* [Monthly min temperature, SSP585](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgD_dataroot/AR6/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/Amon/tasmin/gn/v20190429/tasmin_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc)
* [Monthly max temperature, SSP585](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgD_dataroot/AR6/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/Amon/tasmax/gn/v20190429/tasmax_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc)
* [Monthly mean precipitation, historical](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgC_dataroot/AR6/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/Amon/pr/gn/v20190429/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc)
* [Monthly mean precipitation, SSP585](http://crd-esgf-drc.ec.gc.ca/thredds/fileServer/esgD_dataroot/AR6/CMIP6/ScenarioMIP/CCCma/CanESM5/ssp585/r10i1p1f1/Amon/pr/gn/v20190429/pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc)

* [Land Cover Fraction](http://esgf.nci.org.au/thredds/fileServer/replica/CMIP6/CMIP/CCCma/CanESM5/historical/r10i1p1f1/fx/sftlf/gn/v20190429/sftlf_fx_CanESM5_historical_r10i1p1f1_gn.nc)

## Koeppen-Geiger Climate Classification

Create a Koeppen-Geiger climate classification for the historical period

```{r}
library(terra)
library(climetrics)
library(maps)

# Read data
## Temperature:
tmean <- rast("data/tas_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
tmin <- rast("data/tasmin_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
tmax <- rast("data/tasmax_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")

## Precipitation
pr <- rast("data/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")

## Land Cover Fraction:
lcf <- rast("data/sftlf_fx_CanESM5_historical_r10i1p1f1_gn.nc")

# Get the dates
dates <- time(tmean)
dates <- format(dates, "%Y-%m-%d")
dates <- as.Date(dates)

# Create raster time series
tmean <- rts(tmean, dates)
tmin <- rts(tmin, dates)
tmax <- rts(tmax, dates)
pr <- rts(pr, dates)

# Create a 12-month climatological mean dataset for the 30-year period 1971-01-16 to 2000-12-16
tmean12 <- apply.months(tmean[['1971/2000']],'mean')
tmin12 <- apply.months(tmin[['1971/2000']],'mean')
tmax12 <- apply.months(tmax[['1971/2000']],'mean')
pr12 <- apply.months(pr[['1971/2000']],'mean')

# Convert units from K to C
tmean12 <- tmean12 - 273.15
tmin12 <- tmin12 - 273.15
tmax12 <- tmax12 - 273.15

# Convert units from kg m-2 s-1 to mm per month
pr12 <- pr12 * 86400 * c(31, 28.25, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31) 

# Create a mask that masks out the ocean
## Set all grid cells with a land cover fraction equal to zero to NA:
lcf[lcf == 0] <- NA
## Set all other values equal to one
lcf <- lcf - lcf + 1

k1 <- climetrics::kgc(p=pr12 ,tmin = tmin12 ,tmax=tmax12, tmean = tmean12)

# Mask out the ocean
k1 <- k1 * lcf

# Plot data
my.col <- map.pal("elevation", n = 30)

plot(k1, col = my.col, main = "Koeppen-Geiger Climate Classification \n (CanESM5, historical)")
map("world2", add = TRUE)

```

The numbers in the Figure above refer to the following Koeppen-Geiger climate classification

| Code | Class | Name | Description                                      |
|------|-------|------|--------------------------------------------------|
| 1    | 1     | Af   | Tropical, rainforest                             |
| 2    | 1     | Am   | Tropical, monsoon                                |
| 3    | 1     | Aw   | Tropical, savannah                               |
| 4    | 2     | BWh  | Arid, desert hot                                 |
| 5    | 2     | BWk  | Arid, desert cold                                |
| 6    | 2     | BSh  | Arid, steppe hot                                 |
| 7    | 2     | BSk  | Arid, steppe cold                                |
| 8    | 3     | Csa  | Temperate, dry and hot summer                    |
| 9    | 3     | Csb  | Temperate, dry and warm summer                   |
| 10   | 3     | Csc  | Temperate, dry and cold summer                   |
| 11   | 3     | Cwa  | Temperate, dry winter and hot summer             |
| 12   | 3     | Cwb  | Temperate, dry winter and warm summer            |
| 13   | 3     | Cwc  | Temperate, dry winter and cold summer            |
| 14   | 3     | Cfa  | Temperate, without dry season and hot summer     |
| 15   | 3     | Cfb  | Temperate, without dry season and warm summer    |
| 16   | 3     | Cfc  | Temperate, without dry season and cold summer    |
| 17   | 4     | Dsa  | Cold, dry and hot summer                         |
| 18   | 4     | Dsb  | Cold, dry and warm summer                        |
| 19   | 4     | Dsc  | Cold, dry and cool summer                        |
| 20   | 4     | Dsd  | Cold, dry summer and very cold winter            |
| 21   | 4     | Dwa  | Cold, dry winter and hot summer                  |
| 22   | 4     | Dwb  | Cold, dry winter and warm summer                 |
| 23   | 4     | Dwc  | Cold, dry winter and cool summer                 |
| 24   | 4     | Dwd  | Cold, dry winter and very cold winter            |
| 25   | 4     | Dfa  | Cold, without dry season and hot summer          |
| 26   | 4     | Dfb  | Cold, without dry season and warm summer         |
| 27   | 4     | Dfc  | Cold, without dry season and cool summer         |
| 28   | 4     | Dfd  | Cold, without dry season and very cold winter    |
| 29   | 5     | ET   | Polar, tundra                                    |
| 30   | 5     | EF   | Polar, frost                                     |


Create a Koeppen-Geiger climate classification for the future based on CanESM5 SSP5 for the 2071-2100 period. Call the Koeppen-Geiger climate classification ```k2``` rather than ```k1```. Your output should look similar to the Figure below

```{r, echo = FALSE, eval = TRUE}

library(terra)
library(climetrics)
library(maps)

# Read data
## Temperature:

tmean <- rast("data/tas_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")
tmin <- rast("data/tasmin_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")
tmax <- rast("data/tasmax_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")

## Precipitation
pr <- rast("data/pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")

## Land Cover Fraction:
lcf <- rast("data/sftlf_fx_CanESM5_historical_r10i1p1f1_gn.nc")

# Get the dates
dates <- time(tmean)
dates <- format(dates, "%Y-%m-%d")
dates <- as.Date(dates)

# Create raster time series
tmean <- rts(tmean, dates)
tmin <- rts(tmin, dates)
tmax <- rts(tmax, dates)
pr <- rts(pr, dates)

# Create a 12-month climatological mean dataset for the 30-year period 2071-01-16 to 2100-12-16
tmean12 <- apply.months(tmean[['2071/2100']],'mean')
tmin12 <- apply.months(tmin[['2071/2100']],'mean')
tmax12 <- apply.months(tmax[['2071/2100']],'mean')
pr12 <- apply.months(pr[['2071/2100']],'mean')

# Convert units from K to C
tmean12 <- tmean12 - 273.15
tmin12 <- tmin12 - 273.15
tmax12 <- tmax12 - 273.15

# Convert units from kg m-2 s-1 to mm per month
pr12 <- pr12 * 86400 * c(31, 28.25, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31) 

# Create a mask that masks out the ocean
## Set all grid cells with a land cover fraction equal to zero to NA:
lcf[lcf == 0] <- NA
## Set all other values equal to one
lcf <- lcf - lcf + 1

k2 <- climetrics::kgc(p=pr12 ,tmin = tmin12 ,tmax=tmax12, tmean = tmean12)

# Mask out the ocean
k2 <- k2 * lcf

# Plot data
my.col <- map.pal("elevation", n = 30)

plot(k2, col = my.col, main = "Koeppen-Geiger Climate Classification \n (CanESM5, SSP585)")
map("world2", add = TRUE)

```

Let's plot them next to each other:

```{r}
par(mfrow = c(2,1))

my.col <- map.pal("elevation", n = 30)

plot(k1, col = my.col, main = "Koeppen-Geiger (CanESM5, historical)", cex.main = 0.75)
map("world2", add = TRUE)

plot(k2, col = my.col, main = "Koeppen-Geiger (CanESM5, SSP585)", cex.main = 0.75)
map("world2", add = TRUE)
```

Let's show how the Tundra changes. From the table above you can see that the Tundra corresponds to class ```29```. Isolate class 29 in both ```k1```, and ```k2```, convert them to polygons, and plot them on top of each other using hatched areas.

```{r}

k1.class <- k1
k1.class[k1.class != 29] <- NA
k1.class <- k1.class - k1.class + 1


k2.class <- k2
k2.class[k2.class != 29] <- NA
k2.class <- k2.class - k2.class + 1

# Convert to polygon
k1.class <- as.polygons(k1.class, dissolve = TRUE)
k2.class <- as.polygons(k2.class, dissolve = TRUE)

# Crop region (Extratropics, Northern Hemisphere)
region <- c(0, 360, 40, 90)
k1.class <- crop(k1.class, region)
k2.class <- crop(k2.class, region)

map(
  "world2",
  interior = FALSE,
  boundary = TRUE,
  fill = TRUE,
  col = "grey",
  border = "grey",
  bg = "NA",
  add = FALSE,
  ylim = c(40, 90)
)
box()

plot(
  k1.class,
  density = 15,
  add = TRUE,
  col = "blue",
  border = "blue"
)
plot(
  k2.class,
  density = 15,
  add = TRUE,
  col = "red",
  border = "red",
  angle = -45
)

legend(
  "topleft",
  c("Current Tundra", "Future Tundra") ,
  pch = 16,
  col = c("blue", "red") ,
  bty = "n"
)

```

Repeat the same exercise for a different climate class.

## Velocity of Climate Change

Climate change velocity gives the ratio of the temporal climatic gradient at a given locality to the spatial climatic gradient across neighboring cells.

```{r}

# Read data

tmin.hist <- rast("data/tasmin_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
tmax.hist <- rast("data/tasmax_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")

tmin.ssp <- rast("data/tasmin_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")
tmax.ssp <- rast("data/tasmax_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")

pr.hist <- rast("data/pr_Amon_CanESM5_historical_r10i1p1f1_gn_185001-201412.nc")
pr.ssp <- rast("data/pr_Amon_CanESM5_ssp585_r10i1p1f1_gn_201501-210012.nc")

# Stack past and future together
tmin <- c(tmin.hist, tmin.ssp)
tmax <- c(tmax.hist, tmax.ssp)
pr <- c(pr.hist, pr.ssp)

# Get the dates
dates <- time(tmin)
dates <- format(dates, "%Y-%m-%d")
dates <- as.Date(dates)

# Convert units from K to C
tmin <- tmin - 273.15
tmax <- tmax - 273.15

# Convert units from kg m-2 s-1 to mm per month
days_per_month <- lubridate::days_in_month(dates)
pr <- pr * 86400 * days_per_month

# Create raster time series
tmin <- rts(tmin, dates)
tmax <- rts(tmax, dates)
pr <- rts(pr, dates)

# Mask ocean
lcf <- rast("data/sftlf_fx_CanESM5_historical_r10i1p1f1_gn.nc")
# Create a mask that masks out the ocean
## Set all grid cells with a land cover fraction equal to zero to NA:
lcf[lcf == 0] <- NA
## Set all other values equal to one
lcf <- lcf - lcf + 1
```


```{r}
dv <- climetrics::dVelocity(pr, tmin, tmax, t1='1971/2000',t2='2071/2100')
dv <- dv * lcf

plot(dv, col=map.pal("elevation", n = 100), main='Velocity of Climate Change (km per year)')
map("world2", add = TRUE, interior = FALSE)
```
