---
output:
  pdf_document: default
  html_document: default
---
# Global Earth Observations (II)

In this tutorial ...You will conduct a causal inference analysis that evaluates the statistical associated between satellite data (Enhanved Vegetation Index) and reanalysis (precipitation).



## Download reanalysis data provided by the European Centre for Medium-Range Weather Forecasts (ECMWF)
* Go to the [Copernicus Data Store](https://cds.climate.copernicus.eu/datasets?q=era5&kw=Variable+domain%3A+Atmosphere+%28surface%29&kw=Variable+domain%3A+Atmosphere+%28upper+air%29&kw=Variable+domain%3A+Atmosphere+%28upper+level%29&kw=Variable+domain%3A+Ocean+%28physics%29)
* Click on ```Login-Register```
* Create a ECMWF Account
* After you created your account, return to the data store and select ```ERA5 monthly averaged data on single levels from 1940 to present```
* Click on ```Download```
* Select ```Monthly averaged reanalysis```
* Select "Popular" and "Total Precipitation"
* Select years 2001-2023
* Select all months (Jan - Dec)
* Select the time: 00:00
* Select the Whole available region
* Data format: NetCDF4 (not GRIB!)
* Hit the ```Submit Form``` button
* Wait a minute
* Once your request has been completed, you can hit the download button (379.33 MB)
* Once the download completed, rename your file to ```ERA5_pr_2002-2023.nc``` and move it to your working directory.

## Process Data

### Plot data and verify physicals units

Load the libraries that you will need for this tutorial

```{r, eval = FALSE}
setwd("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430")
.libPaths("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/renv")

library(terra)
library(ncdf4)
library(maps)
library(climetrics)
```

```{r, include = FALSE}
setwd("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430")
.libPaths("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/renv")

library(terra)
library(ncdf4)
library(maps)
library(climetrics)
```

* The variable is the total precipitation measured in m per day.

* Let's plot the annual total precipitation of the first year.

```{r}
rm(list = ls())
data <- terra::rast("data/ERA5_pr_2001-2023.nc")

# Get the first 12 months of data
firstYear <- subset(data, 1:12)

# Calculate the mean and multiply with the number of days per year
firstYear <- mean(firstYear) * 365

# Convert the unit from m to mm
firstYear <- firstYear * 1000

# Define the minimum and maximum values you wish to plot
min_val <- 0
max_val <- 5000

# Define the color breaks
breaks <- seq(min_val, max_val, 250)

my.col <- map.pal("blues", n = length(breaks) - 1)

terra::plot(firstYear, type = "continuous", col = my.col, breaks = breaks, main = "Annual Total Precipitation (mm/yr)")
map("world2", add = TRUE)
```
* Note the white line along the equator?  These are values that exceed 5000 mm per year.

## Spatial resolution

When combining data from different sources, you need to make sure that the data sets use the same spatial grid. Furthermore, you need to mask out all values that both data sets do not have in common. Finally, the analysis you are baout to perform is computational expensive. Choose a relatively coarse grid so that the analysis does not take too long. Run the code below to create a common grid that is a little more coarse than the original one. 

```{r}
rm(list = ls())

# Enhanced Vegetation Index
evi <- terra::rast("MOD13C2061-EVI-0.25deg.nc")
evi <- subset(evi, 12:287) # Subset Jan 2001 to Dec 2023
evi <- aggregate(x = evi, fact=2, fun="mean") # More coarse spatial resolution
mask <- mean(evi) # create mask
mask <- mask - mask + 1 # make boolean mask

# Precipitation
pr <- terra::rast("data/ERA5_pr_2001-2023.nc")
pr <- resample(x = pr, y = evi) # make resolution same as EVI
pr <- pr * mask # exclude gridcells that both data sets do not have in common
```

## Precipitation

* Calculate 12-month climatological means

```{r}
data <- pr

# Create a sequence of dates
start_date <- as.Date("2001-01-15")
end_date <- as.Date("2023-12-15")
dates <- seq(from = start_date, to = end_date, by = "month")

# Create raster time series
data.ts <- rts(data, dates)

# Calculate 12-month climatological means
data.12 <- apply.months(data.ts,'mean')

# Plot 12 months of data
my.col = map.pal("blues", n = 100)
plot(data.12, col = my.col)
```


## Calculate Precipitation Anomalies

```{r}

# Get the number of years
n <- length(dates)/12

# Create a time series where the 12-month climatological mean repeats for all years 
data.clim <- rep(data.12, n)

# Calculate anomalies
data.anom <- data - data.clim

# Have a look at the data

# Set the desired range for the palette
min_val <- -0.03
max_val <- 0.03

# Define breaks to control the min and max of the palette
breaks <- seq(min_val, max_val, 0.005)
my.col <- rev(map.pal("differences", n = length(breaks) - 1))

plot(subset(data.anom, 1:1), type = "continuous", col = my.col, breaks = breaks)
map("world2", add = TRUE)
```
## Detrend precipitation anomalies

```{r}

# Define a function that removes a linear trend
detrend.fun <- function(x) {
  time <- 1:length(x)
  # If a grid cell contains NA, then set the result to NA
  if (is.na(mean(x))) {
    return(rep(NA, length(time)))
  } else {
    time <- 1:length(x)
    linear.model <- lm(x ~ time)
    detrended.series <- stats::residuals(linear.model)
    return(detrended.series)
  }
}

data.anom.detrend <- app(x = data.anom, fun = detrend.fun)

my.col = rev(map.pal("differences", n = 100))
plot(subset(data.anom.detrend, 1:1), col = my.col, main = "Detrended precipitation anomaly (mm/day)")
map("world2", add = TRUE)

pr.anom.detrend <- data.anom.detrend

```
### EVI

Now we need to do the same for the EVI data. We have already done this in the previous tutorial, so I am just adding that same code below in a slightly more concise fashion.

```{r}
data <- evi

# Get the dates
# Create a sequence of dates
start_date <- as.Date("2001-01-15")
end_date <- as.Date("2023-12-15")
dates <- seq(from = start_date, to = end_date, by = "month")

# Create raster time series
data.ts <- rts(data, dates)

# Calculate 12-month climatological means
data.12 <- apply.months(data.ts,'mean')

# Get the number of years
n <- length(dates)/12

# Create a time series where the 12-month climatological mean repeats for all years 
data.clim <- rep(data.12, n)

# Calculate anomalies
data.anom <- data - data.clim

# Define a function that calculates a linear trend
trend.fun <- function(x) {
  
  # If a grid cell contains NA, then set the result to NA
  if (is.na(mean(x))) {
    return(NA)
  } else {
   time <- 1:length(x)
   linear.model <- lm(x ~ time)
   trend <- coef(linear.model)["time"]
  return(trend)
  }
}

data.anom.detrend <- app(x = data.anom, fun = detrend.fun)

plot(subset(data.anom.detrend, 1:1), main = "Detrended EVI anomaly (-)")
map("world2", add = TRUE)

evi.anom.detrend <- data.anom.detrend
```

## Causal inference

Now that you have created detrended anomalies of precipitation and EVI, you can assess the statistical association between both time series using Granger causality. The following analysis is similar to the causal inference analysis we conducted for temperature and albedo for climate model data. Start by defining a Granger causality function that we can apply to each gridcell.

```{r}
granger.fun <- function(x) {
  
    # If a grid cell contains NA, then set the result to NA
  if (is.na(mean(x))) {
    return(NA)
  } else {
    
  n <- length(x)
  # Get first (a) and second (b) variable
  a <- x[1:(n/2)]
  b <- x[(n/2+1):n]
  # Convert to time series
  a <- ts(a)
  b <- ts(b)
  tsDat <- ts.union(a, b)
  tsVAR <- vars::VAR(tsDat, p = 1)
  # Apply Granger causality test 
  p.value <- c(vars::causality(tsVAR, cause = "a")$Granger[3]$p.value)
  return(p.value)
  }
}
```

Check whether precipitation Granger-causes EVI anomalies.

```{r}
# Stack your raster objects, where the first part is the cause and the second the response
data <- c(pr.anom.detrend, evi.anom.detrend)

# Let's focus on Australia
data <- crop(x=data, y = c(110, 160, -45, -10))

# Let's focus on the Amazon basin
# data <- crop(x=data, y = c(285, 315, -15, 5))

# Assess Granger causality  
p.value <- app(x = data, fun = granger.fun)

# Create a boolean map where all locations with p-values < 0.01 equal 1 and all remaining gridcells equal zero
p.value[p.value>=0.05] <- 0
p.value[p.value>0] <- 1
prCausesEvi <- p.value
plot(prCausesEvi, col = c("white", "orange"), main = "Precipitation Granger-causes EVI Anomalies")
map("world2", add = TRUE, interior = FALSE)
```

Check whether EVI Granger-causes precipitation

```{r}
# Stack your raster objects, where the first part is the cause and the second the response
data <- c(evi.anom.detrend, pr.anom.detrend)

# Let's focus on Australia
data <- crop(x=data, y = c(110, 160, -45, -10))

# Let's focus on the Amazon basin
# data <- crop(x=data, y = c(285, 315, -15, 5))

# Assess Granger causality  
p.value <- app(x = data, fun = granger.fun)

# Create a boolean map where all locations with p-values < 0.01 equal 1 and all remaining gridcells equal zero
p.value[p.value>=0.05] <- 0
p.value[p.value>0] <- 1
eviCausesPr <- p.value
plot(eviCausesPr, col = c("white", "orange"), main = "EVI Granger-causes Precipitation Anomalies")
map("world2", add = TRUE, interior = FALSE)
```