# R Basics
* In this course you will be using the programming language R.
* R is a command line driven program. The user enters commands at the prompt and each command is executed one at a time.
* RStudio is a Graphic User Interface for R.
* For this course, you will first need to install R and then RStudio
* If you already have R or RStudio installed, please install them again to ensure you have the most recent version
* Install R from [here](https://posit.co/download/rstudio-desktop)
* Install RStudio from [here](https://posit.co/download/rstudio-desktop)
* Once R and RStudio are installed, open RStudio and learn the basic R syntax following the exercises below.

## Create a new Project
* Open RStudio
* Click on ```File``` in the upper-left corner
* From the drop-down list choose ```New Project ...```, ```New Directory```, ```Browse...``` and choose a directory of your project. For instance, in my case this will be ```C:\Users\seile\Documents\queensu\teaching\classes\ENSC430\tutorials\ENSC430```. Important: make sure your directory does not include white space or special characters.
* Go to your new directory and create a new folder called ```renv```. This is where you will store your packages, as explained next.

## Set up your computational environment
* In the lower-left panel you find your console where you can execute commands. 
* Your first task is to set up your computational environment using the command ```.libPaths()```
* In the Console, type the command below using the directory of your project. Note that you need to use ```/``` rather than ```\```.
```{r, evaluate = FALSE}
# Set your working directory:
setwd("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430")

# Tell R where to store your packages:
.libPaths("C:/Users/seile/Documents/queensu/teaching/classes/ENSC430/tutorials/ENSC430/renv")
```

* You can check if all went well by typing ```.libPaths()```. Does it return your directory? If yes, then continue.
* Next you will install your first package called ```renv```
```{r, eval = FALSE}
install.packages("renv")
renv::init()
```

* Next, install ```rmarkdown```:
```{r, eval = FALSE}
install.packages("rmarkdown")
```

* R Markdown allows you to integrate code, figures, and text. You will write your report in R Markdown.

```{r, eval = FALSE}
install.packages("bookdown")
```

## Syntax

* In R, you use the following syntax to to assign a value to a variable.

```
x <- 1
print(x)
```

* The variable x used above is an example of an R object. The ```print``` statement prints the value of the object x.
* Once you assign values to R objects, you can use the following arithmetic operators to perform calculations.

```
a <- 2
b <- 3

# Addition
c <- a+b
print(c)

# Subtraction
c <- a-b
print(c)

# Multiplication
c <- a*b
print(c)

# Division
c <- a/b
print(c)

# Exponentiation
c <- a^b
print(c)
```
* Note that anything after the # symbol is ignored.

```{r, eval = FALSE}
a <- 2
b <- 3
c <- a + b
# c <- (a + b)^2
print(c)
```
* This way you can add comments into your code that to explain what you are doing

## Objects

* All objects have two intrinsic attributes: mode and length. 
* The mode is the basic type of the elements of the object. 
* There are four main modes: numeric, character, complex, and logical (FALSE or TRUE). 

```{r, eval = FALSE}
a <- 5
class(a)

a <- "apple"
class(a)

a <- 1 + 2i
class(a)

a <- TRUE
class(a)
```
* The length of an object is the number of elements of the object.

```{r, exercise = FALSE, eval = TRUE}
a <- 2
length(a)

a <- c(2,7,5)
length(a)
```
* Next to mode and length, there are different types of objects, namely vector, factor, array, matrix, data frame, and list

```{r, eval = FALSE}
# Vector with one element
a <- 2
print(a)

# Vector with four elements
b <- c(7,5,2,9)
print(b)

# Data frame
x <- c(3,7,2,4)
y <- c(42,67,21,33)
df <- data.frame(x, y)
print(df)

# Matrix: in R, a matrix has 2 dimensions only)
x <- c(1,2,3,4,5,6,7,8,9,10)
a <- matrix(x, nrow = 2)
print(a)

# Array: an array has more than 2 dimensions
x <- c(1,2,3,4,5,6,7,8,9)
a <- array(data = x, dim = c(3,3,3))
print(a)

# List
a <- c(3,2,5,6)
b <- c(TRUE, FALSE)
c <- c("apples", "pears", "plums")
d <- 5

l <- list(a,b,c,d)
print(l)
```

## Generating data

```{r, eval = FALSE}
# Sequence from 1 to 10 with step size 1
x <- 1:10
print(x)

# Sequence from 1 to 10 with step size 0.5
x <- seq(1, 10, 0.5)
print(x)

# Vector that repeats the value one for ten times
x <- rep(1, 10)
print(x)

# Random sequences with uniform distribution between zero and one
x <- runif(n = 5, min = 0, max = 1)
print(x)
```




## Plot data
```{r}
x <- seq(1:10)
y <- x^2
plot(x, y)
```